---
title: "Analysis"
output: html_document
date: "2025-03-02"
---

###Climate-driven variation in the phenology of juvenile Ixodes pacificus on lizard hosts

corresponding author: S. Sambado (sbsambado-at-ucsb.edu) 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary libraries
library(tidyverse)
library(reshape2)
library(mgcv)

## upload necessary data
# lizard-climate
data <- read.csv("data/data_clean_20250302.csv")

# full climate
climate_full_regions <- read.csv("data/climate_full_regions_20250302.csv")
```

#Main Text

#Abstract
```{r}
# range of years
data %>% 
  summarise(year = range(year)) # 2013 - 2022

# how many lizards?
dim(data) # 1527

# how many locations? 
data %>% 
  dplyr::select(location, lat, lon) %>% 
  unique # 45

# what was the range of mean abundance per lizard?
data %>% 
  group_by(location) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE)) %>% 
  arrange(total_mean)
# Coal Oil Point 0.1666667	
# CCSP 47.2058824	

```

#Methods
```{r}

# range of years
data %>% 
  summarise(year = range(year)) # 2013 - 2022


# how many locations? 
data %>% 
  dplyr::select(location, lat, lon) %>% 
  unique # 45


# how many sampling days?
data %>% 
  dplyr::select(location, year, julian) %>% 
  unique # 253

# how many times were locations sampled?
data %>% 
  group_by(location, lat, lon) %>% 
  tally() %>% 
  # filter(n > 1) # 42
  filter(n >= 3) # 38

```

#Results

##Part 1. Characterization of seasonal dynamics

###Field collection summary
```{r}
# how many lizards
dim(data) #1527

# which species of lizards?
data %>% 
  group_by(lizard_species) %>% 
  tally() %>% 
  mutate(prop = n/1527) # calculate proportion of each species


# how many ticks?
data %>% 
  summarise(total = sum(total_ticks, na.rm = TRUE)) # 9338	

# how many ticks identified to larvae
data %>% 
  summarise(total = sum(total_l, na.rm = TRUE)) # 4197	

# how many ticks identified to nymph
data %>% 
  summarise(total = sum(total_n, na.rm = TRUE)) # 3239	


# what proportion were identified?
(3239	+ 4197)/9338 # 0.7963161
```

###Seasonal patterns of juvenile ticks
```{r}
# what was the range of mean abundance per lizard?
data %>% 
  group_by(location) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE)) %>% 
  arrange(total_mean)
# Coal Oil Point 0.1666667	
# CCSP 47.2058824	

# how did these burdens vary by climate region
data %>% 
  filter(climate_region != "North Coast" & climate_region != "San Joaquin Valley") %>% 
  group_by(climate_region) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE),
            sd = sd(total_ticks, na.rm = TRUE)) %>% 
  arrange(desc(total_mean))


## which region had greatest proportion of 0 tick burdens?
data %>% 
    filter(climate_region != "North Coast" & climate_region != "San Joaquin Valley") %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  filter(total_ticks == 0) %>% # select for lizards with 0 burdens
  tally() %>%
  mutate(prop = n/sum(n)) %>% # calc proportions
  arrange(desc(prop))
```

###Phenological Dates
Peak abundance
```{r}
##larva
data %>% 
  filter(month > 2 & month < 7) %>% # select during peak seasonal months
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(total_l = mean(total_l, na.rm = TRUE)) %>% 
  arrange(desc(total_l)) %>% 
  slice(1) %>% 
  # omit regions that are not the main focus of analysis
  filter(climate_region != "North Coast" & climate_region != "San Joaquin Valley") 

##nymph
data %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(total_n = mean(total_n, na.rm = TRUE)) %>% 
  arrange(desc(total_n)) %>% 
  slice(1) %>% 
  # omit regions that are not the main focus of analysis
  filter(climate_region != "North Coast" & climate_region != "San Joaquin Valley") 
```

###Overlap metrics
Kernel Density Estimates (KDE)
```{r}
## Kernal Density Estimates (KDE)

tick_data_kde <- data %>% 
    filter(month > 2 & month < 7) %>% # select during peak seasonal months
    filter(climate_region == "San Francisco Bay Area") %>% # change this out for region
  # filter(climate_region == "Central Coast") %>% # change this out for region
  # filter(climate_region == "Los Angeles") %>% # change this out for region
  group_by(julian) %>% 
  summarise(total_l = mean(total_l, na.rm = TRUE),
            total_n = mean(total_n, na.rm = TRUE)) %>% 
  dplyr::select(julian, total_l, total_n) %>% 
  rename(Date = julian,
         Larvae_Count = total_l,
         Nymphs_Count = total_n) %>% 
  na.omit()

# Create density estimates
larval_density <- density(tick_data_kde$Larvae_Count, from = 0, to = max(tick_data_kde$Larvae_Count))#, n = 512)
nymphal_density <- density(tick_data_kde$Nymphs_Count, from = 0, to = max(tick_data_kde$Nymphs_Count))#, n = 512)

# Calculate overlap area
# After obtaining KDEs for both groups, you can analyze how much they overlap. 
overlap_area <- sum(pmin(larval_density$y, nymphal_density$y)) * (larval_density$x[2] - larval_density$x[1])
overlap_area


# San Francisco Bay Area 0.8881365
# Central Coast 0.3201129
# Los Angeles 0.2953363
```

Jaccard Index (JI)
```{r}
tick_data_ji <- data %>% 
    filter(climate_region == "San Francisco Bay Area") %>% # change this out for region
  # filter(climate_region == "Central Coast") %>% # change this out for region
  # filter(climate_region == "Los Angeles") %>% 
  filter(month > 2 & month < 7) %>% 
  group_by(climate_region, julian) %>%  
  summarise(total_l = mean(total_l, na.rm = TRUE),
            total_n = mean(total_n, na.rm = TRUE)) %>% 
  dplyr::select(julian, total_l, total_n) %>% 
  rename(Date = julian,
         Larvae_Count = total_l,
         Nymphs_Count = total_n) %>% 
  na.omit()

# format data
tick_long <- melt(tick_data_ji, id.vars = "Date", variable.name = "Tick_Type", value.name = "Count")

# calculate
overlap_count <- sum(pmin(tick_data_ji$Larvae_Count, tick_data_ji$Nymphs_Count))
total_count <- sum(pmax(tick_data_ji$Larvae_Count, tick_data_ji$Nymphs_Count))

# Jaccard index
jaccard_index<- overlap_count / total_count
jaccard_index

# SF 0.4725434
# Central Coast  0.3800453
# Los Angeles 0.223288
```


##Part 2
Climate associations with phenological metrics

Format data for models
```{r}
data_3climateregions <- data %>%  
  # mutate NC & SJV to main climate groups
  mutate(climate_region = case_when(climate_region == "North Coast" ~ "San Francisco Bay Area",
                                    climate_region == "San Joaquin Valley" ~ "Los Angeles",
                                    TRUE ~ climate_region)) %>%
  # format climate region N to S
  mutate(climate_region = factor(climate_region,
                                 levels = c("San Francisco Bay Area",
                                            "Central Coast",
                                            "Los Angeles"))) %>% 
  # make sure lat/lon are numeric
  mutate(lat = as.numeric(lat),
         lon = as.numeric(lon))


```

Model 1 - total ticks ~ s(abiotic, by = CR) + CR
CR = climate region (i.e., San Francisco Bay Area (SF), Central Coast (CC), Los Angeles Region (LA))

###All ticks
```{r}
## run model
mod_monthly_3cr <- gam(total_ticks ~ 
               # monthly max temp
             s(tmmx_monthly, by = climate_region, bs = "cr") +
               # monthly specific humidity 
             s(sph_monthly, by = climate_region) +
               # monthly palmer drought severity index
             s(pdsi_monthly, by = climate_region) +
               # account for spatial autocorrelation
              s(lon, lat, bs = "gp") # gaussian process
             + climate_region, # fixed effect
           data = data_3climateregions, 
           method = "REML",
           family = tw(link = "log")) # tweedie distribution

## look at model summary
summary(mod_monthly_3cr)


## check model diagnostics
gam.check(mod_monthly_3cr)

```


###Larvae
```{r}
## run model
mod_monthly_3cr_larvae <- gam(total_l ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
             climate_region, # fixed effect
           data = data_3climateregions, 
           method = "REML",
           family = tw(link = "log"))

## look at model summary
summary(mod_monthly_3cr_larvae)


## check model diagnostics
gam.check(mod_monthly_3cr_larvae)
```


###Nymphs
```{r}
## run model
mod_monthly_3cr_nymphs <- gam(total_n ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
             climate_region, # fixed effect
           data = data_3climateregions, 
           method = "REML",
           family = tw(link = "log"))

## look at model summary
summary(mod_monthly_3cr_nymphs)


## check model diagnostics
gam.check(mod_monthly_3cr_nymphs)

```

