---
title: "4_Analysis"
output: html_document
date: "2024-11-21"
---

Best fit models and analysis
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(mgcv)
```

##Abstract
```{r}
data_post2010 %>% summarise(year = range(year)) # 2013 - 2022

# how many unique locations?
data_post2010 %>% dplyr::select(location,lat, lon) %>% unique # 45


# what did burden range from for locations
data_post2010 %>% 
  group_by(location) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE)) %>% 
  arrange(total_mean)
# Coal Oil Point 0.1666667	
# CCSP 47.2058824	
```


#Methods

```{r}
# range of study
data_post2010 %>% summarise(year = range(year)) # 2013 - 2022

# how many unique locations
data_post2010 %>% dplyr::select(location,lat, lon) %>% unique # 45

# how many days?
data_post2010 %>% dplyr::select(location, year, julian) %>% unique # 253

# how many times were locations sampled
data_post2010 %>% 
  group_by(location,lat, lon) %>%  
  tally() %>% 
  filter(n > 1) # change this number

42/45 # 0.9333333 (more than 1)
38/45 #0.8444444 (more than 3)


# how many lizards
data_post2010 %>% dim() #1527

# how many were western fence lizards
data_post2010 %>%
  group_by(lizard_species) %>% 
  tally() %>% 
  mutate(prop = n/1527) %>%
  mutate_at("prop", round, 3) 
  

# how many ticks per lizard
data_post2010 %>%
  #filter(!grepl("MW", location)) %>%
  #filter(!grepl("CC", location)) %>% 
  summarise(total = sum(total_ticks, na.rm = TRUE)) # 9338

# how many id to life stge
data_post2010 %>%
  summarise(total = sum(total_l, na.rm = TRUE)) # 4197	

data_post2010 %>%
  summarise(total = sum(total_n, na.rm = TRUE)) # 3239	

(4197	+ 3239)/9338 # 0.7963161 ID'd

```

#Results

#Part 1. Characterizing phenology metrics

##A. Density measures 
part a. total ticks 
```{r}
# mean and sd
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE),
            total_sd = sd(total_ticks, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 


# proportion of ticks in march or june
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
  group_by(climate_region, month) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  tally() %>%
  mutate(prop = n/sum(n)) 

# table
# Total
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE),
            total_sd = sd(total_ticks, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 


# Larvae
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_l, na.rm = TRUE),
            total_sd = sd(total_l, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 


# Nymphs
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_n, na.rm = TRUE),
            total_sd = sd(total_n, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 


```

##B. Phenological Dates

part a. Peak abundance
```{r}

##larva
data_post2010_formatted %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(total_l = mean(total_l, na.rm = TRUE)) %>% 
  arrange(desc(total_l)) %>% 
  slice(1)

##nymph
data_post2010_formatted %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(total_n = mean(total_n, na.rm = TRUE)) %>% 
  arrange(desc(total_n)) %>% 
  slice(1)
```

part b. Duration of season
min and max date (mean > 0)
```{r}
## larva
data_post2010_formatted %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(mean_l = mean(total_l, na.rm = TRUE)) %>% 
  filter(mean_l > 0) %>% 
  summarise(min_j = min(julian, na.rm = TRUE),
            max_j = max(julian, na.rm = TRUE),
            length = (max_j-min_j))


## nymph
## larva
data_post2010_formatted %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(mean_n = mean(total_n, na.rm = TRUE)) %>% 
  filter(mean_n > 0) %>% 
  summarise(min_j = min(julian, na.rm = TRUE),
            max_j = max(julian, na.rm = TRUE),
            length = (max_j-min_j))


```

##C. Overlap metrics

part a. KDE
```{r}
tick_data <- data_post2010_formatted %>% 
    filter(month > 2 & month < 7) %>%
    filter(climate_region == "San Francisco Bay Area") %>% # change this out for region
  group_by(julian) %>% 
  summarise(total_l = mean(total_l, na.rm = TRUE),
            total_n = mean(total_n, na.rm = TRUE)) %>% 
  dplyr::select(julian, total_l, total_n) %>% 
  rename(Date = julian,
         Larvae_Count = total_l,
         Nymphs_Count = total_n) %>% 
  na.omit()

# Create density estimates
larval_density <- density(tick_data$Larvae_Count, from = 0, to = max(tick_data$Larvae_Count))#, n = 512)
nymphal_density <- density(tick_data$Nymphs_Count, from = 0, to = max(tick_data$Nymphs_Count))#, n = 512)

# Calculate overlap area
# After obtaining KDEs for both groups, you can analyze how much they overlap. This can be done using various metrics, such as the area under the curves (AUC) or more sophisticated methods that quantify the degree of overlap.
overlap_area <- sum(pmin(larval_density$y, nymphal_density$y)) * (larval_density$x[2] - larval_density$x[1])
overlap_area


# San Francisco Bay Area 0.8881365
# Central Coast 0.3201129
# Los Angeles 0.2953363
```

part b. Jaccard index
```{r}
tick_data <- data_post2010_formatted %>% 
  filter(climate_region == "Los Angeles") %>% 
  filter(month > 2 & month < 7) %>% 
  group_by(climate_region, julian) %>%  
  summarise(total_l = mean(total_l, na.rm = TRUE),
            total_n = mean(total_n, na.rm = TRUE)) %>% 
  dplyr::select(julian, total_l, total_n) %>% 
  rename(Date = julian,
         Larvae_Count = total_l,
         Nymphs_Count = total_n) %>% 
  na.omit()

tick_long <- melt(tick_data, id.vars = "Date", variable.name = "Tick_Type", value.name = "Count")

overlap_count <- sum(pmin(tick_data$Larvae_Count, tick_data$Nymphs_Count))
total_count <- sum(pmax(tick_data$Larvae_Count, tick_data$Nymphs_Count))

# Jaccard index
jaccard_index<- overlap_count / total_count
jaccard_index

# SF 0.4725434
# Central Coast  0.3800453
# Los Angeles 0.223288
```

#Part 2. Climate associations with phenological metrics

Model 1 - total ticks ~ s(abiotic, by = CR) + CR

(All ticks)
```{r}
mod_monthly_by_3cr_mutated <- gam(total_ticks ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
              s(lon, lat, bs = "gp") 
             + climate_region, # fixed effect
           data = master_climate_formatted_scaled_3crs_mutated,
           method = "REML",
           family = tw(link = "log"))

summary(mod_monthly_by_3cr_mutated)

gam.check(mod_monthly_by_3cr_mutated)

## ALl
plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         scale = 0,
         pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])


## TMAX
par(mfrow = c(1,3))
#mod_monthly_tmmx_sf <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly Max Temp (°C)",
         main = "SF",
         select = 1,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_tmmx_sf <- recordPlot()

#mod_monthly_tmmx_cc <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly Max Temp (°C)",
         main = "CC",
         select = 2,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_tmmx_cc <- recordPlot()

#mod_monthly_tmmx_la <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly Max Temp (°C)",
         main = "LA",
         select = 3,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_tmmx_la <- recordPlot()

mod_total_tmmx <- plot_grid(mod_monthly_tmmx_la, align = "hv")

#ggsave(mod_total_tmmx, file = "figures/post2010/mod_total_tmmx.jpeg",
#       height = 3)


## SPH
par(mfrow = c(1,3))
#mod_monthly_tmmx_sf <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly Specific Humidity",
         main = "SF",
         select = 4,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_sph_sf <- recordPlot()

#mod_monthly_tmmx_cc <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly Specific Humidity",
         main = "CC",
         select = 5,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_sph_cc <- recordPlot()

#mod_monthly_tmmx_la <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly Specific Humidity",
         main = "LA",
         select = 6,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_sph_la <- recordPlot()

mod_total_sph <- plot_grid(mod_monthly_sph_la, align = "hv")

#ggsave(mod_total_sph, file = "figures/post2010/mod_total_sph.jpeg",
#       height = 3)


## PDSI
par(mfrow = c(1,3))
#mod_monthly_tmmx_sf <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly PDSI",
         main = "SF",
         select = 7,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_pdsi_sf <- recordPlot()

#mod_monthly_tmmx_cc <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly PDSI",
         main = "CC",
         select = 8,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_pdsi_cc <- recordPlot()

#mod_monthly_tmmx_la <- 
  plot.gam(mod_monthly_by_3cr_mutated,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         xlab = "Monthly PDSI",
         main = "LA",
         select = 9,
         scale = 0,
         #pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated)[1])
#abline(h = 0, col = "red",lwd = 1)
mod_monthly_pdsi_la <- recordPlot()

mod_total_pdsi <- plot_grid(mod_monthly_pdsi_la, align = "hv")

#ggsave(mod_total_pdsi, file = "figures/post2010/mod_total_pdsi.jpeg",
#       height = 3)

```

```{r}
mod_monthly_by_3cr_mutated$sp
```

Larvae
```{r}
mod_monthly_by_3cr_mutated_larva <- gam(total_l ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
              #s(lon, lat, bs = "gp") 
             climate_region, # fixed effect
           data = master_climate_formatted_scaled_3crs_mutated,
           method = "REML",
           family = tw(link = "log"))

summary(mod_monthly_by_3cr_mutated_larva)

gam.check(mod_monthly_by_3cr_mutated_larva)

## ALl
plot.gam(mod_monthly_by_3cr_mutated_larva,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         scale = 0,
         pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated_larva)[1])
mod_monthly_larvae <- recordPlot()

mod_larvae_pdsi <- plot_grid(mod_monthly_larvae, align = "hv")

#ggsave(mod_total_pdsi, file = "figures/post2010/mod_total_pdsi.jpeg",
#       height = 3)


```

Nymphs

```{r}
mod_monthly_by_3cr_mutated_nymph <- gam(total_n ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
              #s(lon, lat, bs = "gp") 
             climate_region, # fixed effect
           data = master_climate_formatted_scaled_3crs_mutated,
           method = "REML",
           family = tw(link = "log"))

summary(mod_monthly_by_3cr_mutated_nymph)

gam.check(mod_monthly_by_3cr_mutated_nymph)

## ALl
plot.gam(mod_monthly_by_3cr_mutated_nymph,
         se = TRUE,
         shade = TRUE, 
         shade.col = "lightblue",
         lwd = 2,
         ylab = "Total Ticks Burden",
         scale = 0,
         pages = 2,
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_by_3cr_mutated_nymph)[1])
mod_monthly_nymph <- recordPlot()

#mod_nymph_pdsi <- plot_grid(mod_monthly_larvae, align = "hv")

#ggsave(mod_total_pdsi, file = "figures/post2010/mod_total_pdsi.jpeg",
#       height = 3)


```
