---
title: "Supplemental Material"
output: html_document
date: "2025-03-02"
---

###Climate-driven variation in the phenology of juvenile Ixodes pacificus on lizard hosts

corresponding author: S. Sambado (sbsambado-at-ucsb.edu) 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(reshape2)
library(mgcv)

## upload necessary data
# lizard-climate
data <- read.csv("data/data_clean_20250302.csv")
# full climate 
climate_full <- read.csv("data/climate_full_clean_20241023.csv") 

## make universal aesthetics
# create universal background layer 
theme_clean <- theme_classic() +
  theme(axis.title = element_text(face = "bold", size = 14),
        axis.text = element_text(size = 14))
# assign colors for each climate region
colors <- c("North Coast" = "#90be6d",
            "San Francisco Bay Area" = "#43aa8b",
            "San Joaquin Valley" = "#f8961e",
            "Central Coast" = "#f9c74f", 
            "Los Angeles" = "#f94144")
```

#Supplemental

##Additional File 1 -- Sampling location and frequency

###Table S1 -- location coordinates
```{r}
table_sup_1_1 <- data %>% 
  dplyr::select(location, collector, lat, lon) %>%
  unique() %>% 
  mutate(across(c(lat:lon), round,5)) %>% 
  arrange(desc(lat))

#write.csv(table_sup_1_1,file = "table_sup_1_1.csv",
#            row.names = FALSE,quote = FALSE)

```


###Table S2 -- location sampling dates
```{r}
table_sup_1_2 <- data %>% 
  filter(month != 9) %>% 
  group_by(location, year, month) %>% 
  tally() %>% 
  pivot_wider(names_from = month, values_from = n)

#write.csv(table_sup_1_2,file = "table_sup_1_2.csv",
#            row.names = FALSE,quote = FALSE)
```


##Additional File 2 -- Location characteristics

###Figure S1 - Sampling frequencies
```{r}
fig_sup_2_1_samplingfrequency <- 
data %>% 
  mutate(location = case_when(location == "Santa Monica Mountains" ~ "SM Mtns",
                              location == "Sedgwick Reserve" ~ "Sedgwick-Sp",
                              location == "Sedgwick" ~ "Sedgwick-Sa",
                              location == "Santa Rosa Island"~ "SR Island",
                              location == "Santa Cruz Island" ~ "SC Island",
                              location == "Paradise Rd unburn" ~ "Paradise Rd",
                              location == "San Luis Obispo" ~ "SLO",
                              TRUE ~ location),
         lat = case_when(location == "Intermediate" ~ 34.97000, # fix so Tejon is next to eachother on y axis
                              TRUE ~ lat)) %>% 
  group_by(location, date, lat, climate_region) %>% 
  tally() %>% 
  ggplot(aes(x = n, y = reorder(location, lat))) +
  geom_bar(stat = "identity", aes(fill = climate_region)) +
  scale_fill_manual(values =  c("North Coast" = "#90be6d",
            "San Francisco Bay Area" = "#43aa8b",
            "San Joaquin Valley" = "#f8961e",
            "Central Coast" = "#f9c74f", 
            "Los Angeles" = "#f94144"), 
            breaks = c("North Coast", "San Francisco Bay Area","San Joaquin Valley",
                       "Central Coast", "Los Angeles")) +
  theme_clean +
  scale_y_discrete(guide = guide_axis(n.dodge = 2)) +
  theme(axis.text.y = element_text(size = 8),
        legend.position = c(.7,.65),
        legend.title = element_text(face = "bold")) +
  labs(x = "Location Visit Frequency",y = "", fill = "Climate Region") +
  geom_vline(xintercept = 3, lty = 2) +
    scale_x_continuous(expand = c(0,0)) +
  annotate(geom = "text", x = 85, y = 31, label = "38°N", size = 6) +
  annotate(geom = "text", x = 85, y = 4, label = "33°N", size = 6)
  
#38.87289
#33.9773

#ggsave(fig_sup_2_1_samplingfrequency, file = "figures/post2010/fig_sup_2_1_samplingfrequency.jpeg", dpi = 600,
#       height = 5, width = 5)


```

###Figure S2 -- lizard captures
```{r}
fig_sup_2_2_lizardcaptures <- 
data %>% 
  mutate(location = case_when(location == "Santa Monica Mountains" ~ "SM Mtns",
                              location == "Sedgwick Reserve" ~ "Sedgwick-Sp",
                              location == "Sedgwick" ~ "Sedgwick-Sa",
                              location == "Santa Rosa Island"~ "SR Island",
                              location == "Santa Cruz Island" ~ "SC Island",
                              location == "Paradise Rd unburn" ~ "Paradise Rd",
                              location == "San Luis Obispo" ~ "SLO",
                              TRUE ~ location),
         lat = case_when(location == "Intermediate" ~ 34.97000, # fix so Tejon is next to eachother on y axis
                              TRUE ~ lat)) %>% 
  group_by(location, lat, climate_region, date) %>% 
  tally() %>% 
  ggplot(aes(x = n, y = reorder(location, lat))) +
  geom_boxplot(aes(fill = climate_region)) +
  scale_fill_manual(values =  c("North Coast" = "#90be6d",
            "San Francisco Bay Area" = "#43aa8b",
            "San Joaquin Valley" = "#f8961e",
            "Central Coast" = "#f9c74f", 
            "Los Angeles" = "#f94144"), 
            breaks = c("North Coast", "San Francisco Bay Area","San Joaquin Valley",
                       "Central Coast", "Los Angeles")) +
  theme_clean +
  scale_y_discrete(guide = guide_axis(n.dodge = 2)) +
  theme(axis.text.y = element_text(size = 8),
        legend.position = "bottom",
        #legend.position = c(.8,.62),
        legend.title = element_text(face = "bold")) +
  labs(x = "Total Lizards Observed",y = "", fill = "")+
    scale_x_continuous(expand = c(0,0)) +
  guides(fill = guide_legend(nrow = 3)) 

#ggsave(fig_sup_2_2_lizardcaptures, file = "figures/post2010/fig_sup_2_2_lizardcaptures.jpeg", 
#       dpi = 600,height = 5, width = 5)
```

###Table S1 -- tick mean and standard deviation
```{r}
table_sup_2_1 <- data %>% 
  group_by(location, lat) %>% 
  summarise(mean_total = mean(total_ticks, na.rm = TRUE),
            sd_total = sd(total_ticks, na.rm = TRUE),
            mean_l = mean(total_l, na.rm = TRUE),
            sd_l = sd(total_l, na.rm = TRUE),
            mean_n = mean(total_n, na.rm = TRUE),
            sd_n = sd(total_n, na.rm = TRUE)) %>% 
  mutate(across(c(mean_total:sd_n), round,3)) %>% 
  arrange(desc(lat))

#write.csv(table_sup_2_1,file = "table_sup_2_1.csv",col.names = FALSE,
#            row.names = FALSE,quote = FALSE)
```

##Additional File 3 -- correlations

###Figure S1 -- Correlation
```{r}

fig_sup_3_1_correlations_numeric <- 
data %>% 
  dplyr::select(tmmn_daily:tmmx_winter) %>% drop_na()

corr <- round(cor(fig_sup_3_1_correlations_numeric),1)

fig_sup_3_1_correlations <- ggcorrplot(corr,
           hc.order = TRUE,
           type = "upper",
           outline.col = "white",
           ggtheme = ggplot2::theme_light(),
           colors = c("#6d9ec1", "white","#e46726"),
           tl.cex = 8)

#ggsave(fig_sup_3_1_correlations, file = "figures/post2010/fig_sup_3_1_correlations.jpeg", dpi = 600,
#       height = 5, width = 5)

```
##Additional File 4 -- Phenological metrics

###Figure S1 -- histogram of juvenile burdens
```{r}


fig_sup_4_1a <- data %>% 
  filter(climate_region == "Central Coast"|
         climate_region == "Los Angeles" |
           climate_region == "San Francisco Bay Area") %>% 
  group_by(location, date, climate_region) %>% 
  summarise(mean_larvae = mean(total_l, na.rm = TRUE)) %>% 
  ggplot(aes(x = mean_larvae))+
  geom_histogram(fill = "grey95", color = "grey15") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_clean +
  facet_wrap(~factor(climate_region, c("San Francisco Bay Area",
                                       "Central Coast",
                                       "Los Angeles")), ncol = 1) +
  theme_minimal() +
  theme(strip.text = element_text(face= "bold", size = 12)) +
  labs(x = "Mean Larvae on Individual Lizard", y = "Frequency",
       tags = "A")


fig_sup_4_1b <- data %>% 
  filter(climate_region == "Central Coast"|
         climate_region == "Los Angeles" |
           climate_region == "San Francisco Bay Area") %>% 
  group_by(location, date, climate_region) %>% 
  summarise(mean_nymphs = mean(total_n, na.rm = TRUE)) %>% 
  ggplot(aes(x = mean_nymphs))+
  geom_histogram(fill = "grey95", color = "grey15") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_clean +
  facet_wrap(~factor(climate_region, c("San Francisco Bay Area",
                                       "Central Coast",
                                       "Los Angeles")), ncol = 1) +
  theme_minimal() +
  theme(strip.text = element_text(face= "bold", size = 12)) +
  labs(x = "Mean Nymphs on Individual Lizard", y = "Frequency",
       tags = "B")


fig_sup_4_1 <- ggarrange(fig_sup_4_1a ,fig_sup_4_1b)

#ggsave(fig_sup_4_1, file = "figures/post2010/fig_sup_4_1.jpeg", dpi = 600,
#       height = 5, width = 7)

```


###Figure S2 -- juvenile burdens by month and year
```{r}
fig_sup_4_1a <- data %>% 
  filter(month != 9) %>% 
  group_by(climate_region,location, date, month) %>% 
  ggplot(aes(x = as.factor(month), y = total_ticks))  +
  scale_color_manual(values =  c("North Coast" = "#90be6d",
            "San Francisco Bay Area" = "#43aa8b",
            "San Joaquin Valley" = "#f8961e",
            "Central Coast" = "#f9c74f", 
            "Los Angeles" = "#f94144")) +
    geom_jitter(aes(color = climate_region), alpha = .5, width = .25) +
    geom_boxplot(fill = "grey90") +
  theme_clean +
  scale_x_discrete(labels = c("Feb", "March", "April", 
                              "May", "June")) +
  labs(x = "", y = "Raw Tick Counts per Lizard", color = "", tag = "A") +
  #guides(color = FALSE)
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold")) +
  guides(color = guide_legend(nrow = 1) ) 


fig_sup_4_1b <- data %>% 
  group_by(climate_region,location, date, year) %>% 
  ggplot(aes(x = as.factor(year), y = total_ticks)) +
  scale_color_manual(values =  c("North Coast" = "#90be6d",
            "San Francisco Bay Area" = "#43aa8b",
            "San Joaquin Valley" = "#f8961e",
            "Central Coast" = "#f9c74f", 
            "Los Angeles" = "#f94144")) +
    geom_jitter(aes(color = climate_region), alpha = .5, width = .25) +
    geom_boxplot(fill = "grey90") +
  theme_clean +
  labs(x = "", y = "Raw Tick Counts per Lizard", color = "", tag = "B") +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold")) +
  guides(color = guide_legend(nrow = 1) )

fig_sup_4_1 <- ggarrange(fig_sup_4_1a, fig_sup_4_1b + rremove("ylab"),
          common.legend = TRUE, ncol = 2,
          legend = "bottom",
          align = "hv", widths = c(.8,1))

#ggsave(fig_sup_4_1, file = "figures/post2010/fig_sup_4_1.jpeg", dpi = 600,
#       height = 5, width = 10)

```
###Table S1 -- Phenology metrics

Density
```{r}
#table_sup_2_1 <- data

# Total
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_ticks, na.rm = TRUE),
            total_sd = sd(total_ticks, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 


# Larvae
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_l, na.rm = TRUE),
            total_sd = sd(total_l, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 


# Nymphs
data_post2010 %>% 
  group_by(climate_region) %>% 
  filter(month > 2 & month < 7) %>% 
    mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  summarise(total_mean = mean(total_n, na.rm = TRUE),
            total_sd = sd(total_n, na.rm = TRUE)) %>% 
  mutate_at(c("total_mean","total_sd"), round, 2) 

```


Peak abundance
```{r}

##larva
data %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(total_l = mean(total_l, na.rm = TRUE)) %>% 
  arrange(desc(total_l)) %>% 
  slice(1)

##nymph
data %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(total_n = mean(total_n, na.rm = TRUE)) %>% 
  arrange(desc(total_n)) %>% 
  slice(1)
```

Duration of season
min and max date (mean > 0)
```{r}
## larva
data %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(mean_l = mean(total_l, na.rm = TRUE)) %>% 
  filter(mean_l > 0) %>% 
  summarise(min_j = min(julian, na.rm = TRUE),
            max_j = max(julian, na.rm = TRUE),
            length = (max_j-min_j))


## nymph
## larva
data %>% 
  filter(month > 2 & month < 7) %>% 
  mutate(climate_region = factor(climate_region,
                                 levels = c("North Coast", 
                                   "San Francisco Bay Area", "San Joaquin Valley",
                                   "Central Coast", "Los Angeles"))) %>% 
  group_by(climate_region, julian) %>% 
  summarise(mean_n = mean(total_n, na.rm = TRUE)) %>% 
  filter(mean_n > 0) %>% 
  summarise(min_j = min(julian, na.rm = TRUE),
            max_j = max(julian, na.rm = TRUE),
            length = (max_j-min_j))


```

Overlap metrics

part a. KDE
```{r}
tick_data_kde <- data %>% 
    filter(month > 2 & month < 7) %>%
    filter(climate_region == "San Francisco Bay Area") %>% # change this out for region
  group_by(julian) %>% 
  summarise(total_l = mean(total_l, na.rm = TRUE),
            total_n = mean(total_n, na.rm = TRUE)) %>% 
  dplyr::select(julian, total_l, total_n) %>% 
  rename(Date = julian,
         Larvae_Count = total_l,
         Nymphs_Count = total_n) %>% 
  na.omit()

# Create density estimates
larval_density <- density(tick_data_kde$Larvae_Count, from = 0, to = max(tick_data_kde$Larvae_Count))#, n = 512)
nymphal_density <- density(tick_data_kde$Nymphs_Count, from = 0, to = max(tick_data_kde$Nymphs_Count))#, n = 512)

# Calculate overlap area
# After obtaining KDEs for both groups, you can analyze how much they overlap. This can be done using various metrics, such as the area under the curves (AUC) or more sophisticated methods that quantify the degree of overlap.
overlap_area <- sum(pmin(larval_density$y, nymphal_density$y)) * (larval_density$x[2] - larval_density$x[1])
overlap_area


# San Francisco Bay Area 0.8881365
# Central Coast 0.3201129
# Los Angeles 0.2953363
```

part b. Jaccard index
```{r}
tick_data_ji <- data %>% 
  filter(climate_region == "Los Angeles") %>% 
  filter(month > 2 & month < 7) %>% 
  group_by(climate_region, julian) %>%  
  summarise(total_l = mean(total_l, na.rm = TRUE),
            total_n = mean(total_n, na.rm = TRUE)) %>% 
  dplyr::select(julian, total_l, total_n) %>% 
  rename(Date = julian,
         Larvae_Count = total_l,
         Nymphs_Count = total_n) %>% 
  na.omit()

tick_long <- melt(tick_data_ji, id.vars = "Date", variable.name = "Tick_Type", value.name = "Count")

overlap_count <- sum(pmin(tick_data_ji$Larvae_Count, tick_data_ji$Nymphs_Count))
total_count <- sum(pmax(tick_data_ji$Larvae_Count, tick_data_ji$Nymphs_Count))

# Jaccard index
jaccard_index<- overlap_count / total_count
jaccard_index

# SF 0.4725434
# Central Coast  0.3800453
# Los Angeles 0.223288
```

##Additional File 5 -- GAMs

###All ticks
```{r}
## run model
mod_monthly_3cr <- gam(total_ticks ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
              s(lon, lat, bs = "gp") 
             + climate_region, # fixed effect
           data = data_3climateregions, 
           method = "REML",
           family = tw(link = "log"))

## look at model summary
summary(mod_monthly_3cr)


## check model diagnostics
gam.check(mod_monthly_3cr)

## Plot model results for all ticks
plot.gam(mod_monthly_3cr,
         se = TRUE, # leave standard error
         shade = TRUE, # shade ci
         shade.col = "lightblue",
         lwd = 2, # line thickness
         ylab = "Juvenile Tick Burden", 
         scale = 0,
         pages = 2, # put on 2 pages
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_3cr)[1]) # shift y intercept based on coeficients
```

###Larvae
```{r}
## run model
mod_monthly_3cr_larvae <- gam(total_l ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
             climate_region, # fixed effect
           data = data_3climateregions, 
           method = "REML",
           family = tw(link = "log"))

## look at model summary
summary(mod_monthly_3cr_larvae)


## check model diagnostics
gam.check(mod_monthly_3cr_larvae)

## Plot model results for larval ticks
plot.gam(mod_monthly_3cr_larvae,
         se = TRUE, # leave standard error
         shade = TRUE, # shade ci
         shade.col = "lightblue",
         lwd = 2, # line thickness
         ylab = "Larval Tick Burden", 
         scale = 0,
         pages = 2, # put on 2 pages
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_3cr_larvae)[1]) # shift y intercept based on coeficients
mod_monthly_larvae <- recordPlot()

#mod_larvae_pdsi <- plot_grid(mod_monthly_larvae, align = "hv")

#ggsave(mod_monthly_larvae, file = "figures/post2010/mod_monthly_larvae.jpeg",
#       height = 3)
```

###Nymphs
```{r}
## run model
mod_monthly_3cr_nymphs <- gam(total_n ~ 
             s(tmmx_monthly, by = climate_region, bs = "cr") +
             s(sph_monthly, by = climate_region) +
             s(pdsi_monthly, by = climate_region) +
             climate_region, # fixed effect
           data = data_3climateregions, 
           method = "REML",
           family = tw(link = "log"))

## look at model summary
summary(mod_monthly_3cr_nymphs)


## check model diagnostics
gam.check(mod_monthly_3cr_nymphs)

## Plot model results for larval ticks
plot.gam(mod_monthly_3cr_nymphs,
         se = TRUE, # leave standard error
         shade = TRUE, # shade ci
         shade.col = "lightblue",
         lwd = 2, # line thickness
         ylab = "Nymphal Tick Burden", 
         scale = 0,
         pages = 2, # put on 2 pages
         scheme = 2,
         seWithMean = TRUE,
         shift = coef(mod_monthly_3cr_nymphs)[1]) # shift y intercept based on coeficients
mod_monthly_nymphs <- recordPlot()

mod_nymphs <- plot_grid(mod_monthly_nymphs, align = "hv")


#ggsave(mod_nymphs, file = "figures/post2010/mod_monthly_nymphs.jpeg",
#       height = 3)
```